---
title: "Re-development of clinical trials of Imatinib"
author: "Yongxi Long"
format: 
  html:
    self-contained: true
toc: true
editor: visual
---

## Aim

Evaluate the performance of different dose-finding methods on Imatinib profile constructed from trial reports. The Phase II and Phase III design used in the simulation study are different from the actual trial designs of Imatinib, to accommodate the fact that only one dose will be carried forward in the simulation study.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsci)
  library(RColorBrewer)
  library(trialr)
  library(clinfun)
  library(ggExtra)
  library(ggridges)
  library(ggpubr)
})
source("functions/combine_res.R")
# load data
load("Imatinib_example/Imatinib_data/Imatinib_data.RData")
```

## Phase I

The sample size is 40 for the actual Phase I study

```{r}
N_ph1 <- 40
nsim <- 1e3
# target DLT for model-assisted and model-based design
target_DLT <- 0.3
```

### 3+3 design

```{r,eval=FALSE}
source("functions/threep3.R")
res_ph1_threep3 <- sapply(Imatinib_data, function(data)
  {
  run_threep3(data = data,
              N=N_ph1,
              drug_detail = FALSE)
},simplify = FALSE)
combine_res(res_ph1 = res_ph1_threep3)
save(res_ph1_threep3,
     file = paste0("Imatinib_example/Imatinib_simres/res_ph1_threep3.RData"))
```

### BOIN design

```{r}
source("functions/BOIN.R")
# cohort size
cohort_size= 3
# toxicity bound for inadmissible doses
tox_bound <- 0.95
# BOIN design
res_ph1_BOIN <- sapply(Imatinib_data, function(data)
  {
  run_BOIN(data = data,
                 N=N_ph1,
                 cohort_size = cohort_size,
                 target_DLT=target_DLT,
                 tox_bound = tox_bound,
                drug_detail = FALSE)
},simplify = FALSE)
# the for loop is for debugging
# res_ph1_BOIN <- vector(mode = "list",length = nsim)
# for(i in 1:nsim)
# {
# res <- run_BOIN(data = Imatinib_data[[i]],
#                  N=N_ph1,
#                  cohort_size = cohort_size,
#                  target_DLT=target_DLT,
#                  tox_bound = tox_bound,
#                 drug_detail = FALSE)
# res_ph1_BOIN[[i]] <- res
# }
combine_res(res_ph1 = res_ph1_BOIN)
save(res_ph1_BOIN,
     file = paste0("Imatinib_example/Imatinib_simres/res_ph1_BOIN.RData"))

```

### CRM Design

```{r}
source("functions/CRM.R")
# cohort size
cohort_size= 1
# toxicity boundary
tox_bound <- 0.9
# from paper Chu et. al
# Unifying CRM and EWOC designs for Phase I cancer clinical trials
res_ph1_CRM <- vector(mode = "list",length = nsim)
for(i in 1:nsim)
{
  res <- run_CRM(data=Imatinib_data[[i]],
                    N=N_ph1,
                    cohort_size = cohort_size,
                    target_DLT=target_DLT,
                    tox_bound = tox_bound,
                    model="empiric",
                 drug_detail = FALSE)
  res_ph1_CRM[[i]] <- res
  print(i)
}
save(res_ph1_CRM,file = paste0("Imatinib_example/Imatinib_simres/res_ph1_CRM.RData"))
combine_res(res_ph1 = res_ph1_CRM)
```
