---
title: "Sensitivity analysis for BOIN12"
author: "Yongxi Long"
format: 
  html:
    self-contained: true
editor: visual
---

## Aim

BOIN12 involves user-specified utility values for four pairs of outcome combinations: $u1$ for (Efficacy, No toxicity), $u2$ for (No efficacy, No Toxicity), $u3$ for (Efficacy, Toxicity), and $u4$ for (No efficacy, Toxicity). Usually $u1$ is set to 100 and $u4$ is set to 0 to benchmark the most/least desirable outcome scenario. The relative value of $u2$ and $u3$ reflects clinical considerations about the trade-off between efficacy and toxicity. If $u2 > u3$, we weigh toxicity more than efficacy.

This script is to perform sensitivity analysis on the choice of the utility values for $u2$ and $u3$. For simplicity we constrain that $u2+u3=100$ such that there is only one tuning variable. We assess how various values of $u2$ affects the final probability of success across the three drug families.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsci)
  library(RColorBrewer)
  library(trialr)
  library(clinfun)
  library(ggExtra)
  library(ggridges)
  library(ggpubr)
})
source("functions/combine_res.R")
```

## Simulation for a range of utility values

```{r,echo=FALSE}
# set up simulation scenarios
source("functions/BOIN12.R")
# the toxicity upper limit
phi_t <- 0.35
# the efficacy lower limit
phi_e <- 0.5
# when tox is within bounds, 
# stop exploring higher doses when n at dose is greater than or equal to this value.
n_star1 <- 6
# dose-exploration rule, go to the next higher dose if more than this number of patients have been treated at current dose and the current dose is safe enough
# and the next higher dose has not yet been explored
n_star2 <- 8
# combinations of utility values
# utility for outcome no tox no response
u2s <- seq(10,90,by=10)
# utility for outcome tox yes and response yes
u3s <- rev(u2s)
# drug families
drug_families <- c("immuno_fixed_OBD3_MTD5", "immuno_fixed_OBD4_MTD5", "chemo_fixed_MTD5" )
```

```{r,echo=FALSE,eval=FALSE}
# run the simulation
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
for(sce in 1:length(u2s))
{
u2 <- u2s[sce]; u3 <- u3s[sce]
res_ph1_BOIN12 <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  dat <- drug_data[[i]]
  res <- suppressWarnings({
    run_BOIN12(data = dat,
               N= N,
               cohort_size = 3,
               phi_t = phi_t, phi_e = phi_e,
               u2 = u2, u3 = u3,
               n_star1 = n_star1, n_star2 = n_star2)
  })
  res_ph1_BOIN12[[i]] <- res
  # print(i)
}
save(res_ph1_BOIN12,file = paste0("results/res_BOIN12/res_ph1_",
                                  drug_family,
                                  "_BOIN12_sce",
                                  sce,".RData"))
print(paste0(drug_family,": scenario",sce))
}
}
```

## Visualize results

```{r,echo=FALSE,warning=FALSE}
drug_family <- drug_families[3]
RP2D_summary_OBD <- purrr::map_dfr(1:9,function(sce)
{
  load(paste0("results/res_BOIN12/res_ph1_",
              drug_family,
              "_BOIN12_sce",
              sce,".RData"))
  # Retrieve the actual object (e.g., res_ph1_threep3)
  res_ph1_obj <- res_ph1_BOIN12
  res <- combine_res(res_ph1 = res_ph1_obj)
  res <- res$`Table of RP2D` |>
    select(dose_level,prob_selected) |>
    #mutate(Design = design) |>
    mutate(utility2 = u2s[sce], Utility3 = u3s[sce]) 
    # mutate(Design = factor(Design,levels=c("threep3","BOIN","CRM","EffTox"),
    #                        labels = c("Threep3","BOIN","CRM","EffTox"))) |>
    # mutate(Drug_family = case_when(
    #   drug_family == "immuno_fixed_OBD3_MTD5" ~ "immuno OBD=3",
    #   drug_family == "immuno_fixed_OBD4_MTD5" ~ "immuno OBD=4",
    #   drug_family == "chemo_fixed_MTD5" ~ "chemo MTD=5"
    # ))
  return(res)
})

drug_family_label <- case_when(
      drug_family == "immuno_fixed_OBD3_MTD5" ~ "immuno-like profile OBD=3",
      drug_family == "immuno_fixed_OBD4_MTD5" ~ "immuno-like profile OBD=4",
      drug_family == "chemo_fixed_MTD5" ~ "chemo-like profile MTD=5"
    )
# barplot 
# RP2D_summary_OBD |>
#   ggplot(aes(x = factor(utility2), y = prob_selected, fill = factor(dose_level))) +
#   geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
#   theme_bw() +
#   labs(
#     x = "u2 (u3=100-u2)",
#     y = "Probability of Selection",
#     fill = "RP2D",
#     title = drug_family_label
#   )+
#   geom_text(aes(label = round(prob_selected, 2)),  # text labels
#             position = position_dodge(width = 0.8), 
#             vjust = -0.3,                          # vertical position (above bar)
#             size = 3,
#             angle = 45)+
#   scale_fill_viridis_d(option = "viridis",na.value="grey80")+
#   theme(
#     title = element_text(size=9)
#   )+
#   ylim(0,0.7)


RP2D_summary_OBD |>
  ggplot(aes(x = factor(utility2), y = prob_selected, group = factor(dose_level),
             color=factor(dose_level))) +
  geom_line(lwd=0.7)+
  geom_point()+
  theme_bw() +
  labs(
    x = "u2 (u3=100-u2)",
    y = "Probability of Selection",
    color = "RP2D",
    title = drug_family_label
  )+
  geom_text(aes(label = round(prob_selected, 2)),  # text labels
            position = position_dodge(width = 0.8), 
            vjust = -0.3,                          # vertical position (above bar)
            size = 3,
            angle = 45,
            show.legend = FALSE)+
  scale_color_viridis_d(option = "viridis",na.value="grey80")+
  theme(
    title = element_text(size=9)
  )+
  ylim(0,0.7)

```

## Combine Phase I-III results

```{r}
u2s <- seq(10,90,by=10)
# utility for outcome tox yes and response yes
u3s <- rev(u2s)
#drug_family <- "chemo_fixed_MTD5"
# drug_family <- "immuno_fixed_OBD4_MTD5"
# drug_family <- "immuno_fixed_OBD3_MTD5"
p_list <- sapply(drug_families, function(drug_family)
{
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
res_ph123_df <- purrr::map_dfr(TS_rules,function(TS)
  {
  load(paste0("results/res_ph2_",drug_family,"_simon2stage_TS",TS,".RData"))
  load(paste0("results/res_ph3_",drug_family,"_RCT_TS",TS,".RData"))
  out <- purrr::map_dfr(1:9,function(sce)
{
  load(paste0("results/res_BOIN12/res_ph1_",
              drug_family,
              "_BOIN12_sce",
              sce,".RData"))
    res_ph1_obj <- res_ph1_BOIN12
    res <- combine_res(res_ph1 = res_ph1_obj,
                       res_ph2 = res_ph2_simon2stage,
                       res_ph3 = res_ph3_RCT)
    out <- data.frame(
      Ph123_PoS = res$`Phase I-III probability of significance`,
      Ph23_PoTS = res$`Phase II&III probability of toxicity suspension`
    ) |>
      mutate(u2 = u2s[sce],u3=u3s[sce])
    return(out)
  })
  out <- out |>
    mutate( TS_rule = TS
    )
  return(out)
})
res_ph123_df <- res_ph123_df |>
  mutate(TS_rule = case_when(
    is.na(TS_rule) ~ "None",
    TRUE ~ paste0(TS_rule*100,"% DLT")
  ))

drug_family_label <- case_when(
      drug_family == "immuno_fixed_OBD3_MTD5" ~ "immuno-like profile OBD=3",
      drug_family == "immuno_fixed_OBD4_MTD5" ~ "immuno-like profile OBD=4",
      drug_family == "chemo_fixed_MTD5" ~ "chemo-like profile MTD=5"
    )
p <- res_ph123_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph123_PoS,color=factor(u2),
             group = factor(u2)))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Phase I-III Probability of Significance (PoS)",
       title = drug_family_label,
       color = "Utility for (No efficacy, No toxicity)")+
  # scale_color_viridis_d()+
  scale_color_manual(values = c("#440154FF", "#46337EFF", "#365C8DFF", "red",
                                "#277F8EFF", "#1FA187FF", "#4AC16DFF" ,"#9FDA3AFF", "#FDE725FF"))+
  theme(
    legend.position = "right"
  )
  return(p)
},simplify = FALSE)
# immuno-like
ggarrange(plotlist = p_list[1:2],ncol = 2,
          common.legend = TRUE)
# chemo-like
p_list[[3]]
```
