---
title: "Sensitivity analysis of EffTox"
author: "Yongxi Long"
format: 
  html:
      self-contained: true
code-fold: true
editor: visual
---

## Aim

EffTox design involves multiple user-speficied tuning parameters, some are usually set to certain standard values or have a clear clinical reference, for example, the efficacy hurdle is often the desired response rate at later Phase II and the toxicity hurdle the target DLT. Some, however, are more ambigous to specify from a clinical perspective, such as the three points defining the Contour plot.

This script will investigate the impact of the contour plot specification on the performance of EffTox across the three drug families

```{r,echo=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsci)
  library(RColorBrewer)
  library(trialr)
  library(clinfun)
  library(ggExtra)
  library(ggridges)
  library(ggpubr)
})
source("functions/combine_res.R")
```

## Simulation for a series of Contour plots

```{r,echo=FALSE}
source("functions/EffTox.R")
# trial design parameters
# sample size
N <- 30
# cohort size
cohort_size <- 3
# Threshold for selecting desirable doses
efficacy_hurdle <- 0.5 # desired response rate
toxicity_hurdle <- 0.3 # target DLT
p_e <- p_t <- 0.05
# prior guess of DLT/RSP prob for each dose
pi_T <- dfcrm::getprior(halfwidth = 0.05,target = 0.3,nu=3,nlevel=6,model="empiric")
pi_E <- dfcrm::getprior(halfwidth = 0.05,target = 0.5,nu=3,nlevel=6,model="empiric")
# effective sample size for the prior
ess_T <- ess_E <- 1
# get normal prior hyperparameters for the EffTox model
# it can take some time for iterative calculation
# the dose levels are already scaled between 0-1, it does 
# not impact the prior hyperparamters that much
# so for saving time, will use the same (average) actual dose
# levels for all simulation iteration, although the actual dose
# levels are different for each iteration
priors_EffTox <- get_efftox_priors(doses = c(0.1000000, 
                                             0.2000000, 
                                             0.3300000, 
                                             0.5016000, 
                                             0.7022400, 
                                             0.9339792),
                                   pi_T = pi_T,
                                   ess_T = ess_T,
                                   pi_E = pi_E,
                                   ess_E = ess_E)
# the Contour
# smallest rsp acceptable if there is no toxicity, set to around the SOC, p0 = 0.4
eff0_vec <- c(0.35,0.45)
# highest DLT acceptable if the rsp is 100%, set to slightly more than the target DLT 30%
tox1_vec <- c(0.5,0.8)
sce_df <- expand.grid(eff0=eff0_vec,tox1=tox1_vec)
sce_df <- rbind(sce_df,sce_df)
# middle neutral utility value
# slight more toxic than target, and slight less efficacious than target
sce_df$eff_star <- c(rep(0.65,4),rep(0.55,4))
sce_df$tox_star <- c(rep(0.25,4),rep(0.45,4))
main_sce_df <- data.frame(
  eff0 = 0.4,
  tox1 = 0.7,
  eff_star = 0.65,
  tox_star = 0.25
)
# drug families
drug_families <- c("immuno_fixed_OBD3_MTD5", "immuno_fixed_OBD4_MTD5", "chemo_fixed_MTD5" )
```

```{r,echo=FALSE,eval=FALSE}
all_sce_df <- rbind(main_sce_df,sce_df)
contour_plots <- vector(mode = "list",length = nrow(all_sce_df))
for(i in 1:nrow(all_sce_df))
{
  eff0 <- all_sce_df$eff0[i]; tox1 <- all_sce_df$tox1[i]
  eff_star <- all_sce_df$eff_star[i]; tox_star <- all_sce_df$tox_star[i]
  p_EffTox <-  trialr::efftox_solve_p(eff0,tox1,eff_star,tox_star)

  fit_EffTox <- stan_efftox(
        outcome_str = "1NNN",
        real_doses = drug_data[[1]]$profile_info$dose,
        efficacy_hurdle = efficacy_hurdle,
        toxicity_hurdle = toxicity_hurdle,
        p_e = p_e,
        p_t = p_t,
        eff0 = eff0,
        tox1 = tox1,
        eff_star = eff_star,
        tox_star = tox_star,
        priors = priors_EffTox,
        refresh = 0
      )
  contour_plots[[i]] <- efftox_contour_plot(fit_EffTox,use_ggplot = TRUE)+
    theme(
      axis.title = element_text(size=8)
    )
}
save(contour_plots,file = "results/res_EffTox/Contour_plots.RData")
```

```{r,echo=FALSE,warning=FALSE}
load("results/res_EffTox/Contour_plots.RData")
ggarrange(plotlist = contour_plots,nrow=3,ncol=3)

ggarrange(plotlist = contour_plots[2:9],nrow=2,ncol=4,
          labels = 1:8)
```



```{r,eval=FALSE,echo=FALSE}
# run the simulation
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
for(sce in 1:nrow(sce_df))
{
eff0 <- sce_df$eff0[sce]; tox1 <- sce_df$tox1[sce]
eff_star <- sce_df$eff_star[sce]; tox_star <- sce_df$tox_star[sce]
p_EffTox <-  trialr::efftox_solve_p(eff0,tox1,eff_star,tox_star)
res_ph1_EffTox <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  dat <- drug_data[[i]]
  res <- suppressWarnings({
    run_EffTox(
      data = dat,
      N=N,
      cohort_size=cohort_size,
      efficacy_hurdle = efficacy_hurdle, 
      toxicity_hurdle = toxicity_hurdle,
      p_e = p_e, p_t = p_t,
      eff0 = eff0, tox1 = tox1,
      eff_star = eff_star, tox_star = tox_star,
      priors = priors_EffTox)
  })
  res_ph1_EffTox[[i]] <- res
  print(i)
}
save(res_ph1_EffTox,file = paste0("results/res_EffTox/res_ph1_",
                                  drug_family,
                                  "_EffTox_sce",
                                  sce,".RData"))
}
  }
```

## Visualize results

```{r,echo=FALSE,warning=FALSE}
drug_family <- drug_families[2]
RP2D_summary_OBD <- purrr::map_dfr(1:nrow(sce_df),function(sce)
{
  load(paste0("results/res_EffTox/res_ph1_",
              drug_family,
              "_EffTox_sce",
              sce,".RData"))
  # Retrieve the actual object (e.g., res_ph1_threep3)
  res_ph1_obj <- res_ph1_EffTox
  res <- combine_res(res_ph1 = res_ph1_obj)
  res <- res$`Table of RP2D` |>
    select(dose_level,prob_selected) |>
    mutate(Scenario = sce) 
    # mutate(Design = factor(Design,levels=c("threep3","BOIN","CRM","EffTox"),
    #                        labels = c("Threep3","BOIN","CRM","EffTox"))) |>
  return(res)
})

drug_family_label <- case_when(
      drug_family == "immuno_fixed_OBD3_MTD5" ~ "immuno-like profile OBD=3",
      drug_family == "immuno_fixed_OBD4_MTD5" ~ "immuno-like profile OBD=4",
      drug_family == "chemo_fixed_MTD5" ~ "chemo-like profile MTD=5"
)


RP2D_summary_OBD |>
  ggplot(aes(x = factor(Scenario), y = prob_selected, group = factor(dose_level),
             color=factor(dose_level))) +
  geom_line(lwd=0.7)+
  geom_point()+
  theme_bw() +
  labs(
    x = "Contour setting",
    y = "Probability of Selection",
    color = "RP2D",
    title = drug_family_label
  )+
  geom_text(aes(label = round(prob_selected, 2)),  # text labels
            position = position_dodge(width = 0.8), 
            vjust = -0.3,                          # vertical position (above bar)
            size = 3,
            angle = 45,
            show.legend = FALSE)+
  scale_color_viridis_d(option = "viridis",na.value="grey80")+
  theme(
    title = element_text(size=9)
  )+
  ylim(0,0.7)

```


## Combine Phase I-III results

```{r,echo=FALSE}
#drug_family <- "chemo_fixed_MTD5"
# drug_family <- "immuno_fixed_OBD4_MTD5"
# drug_family <- "immuno_fixed_OBD3_MTD5"
p_list <- sapply(drug_families, function(drug_family)
{
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
res_ph123_df <- purrr::map_dfr(TS_rules,function(TS)
  {
  load(paste0("results/res_ph2_",drug_family,"_simon2stage_TS",TS,".RData"))
  load(paste0("results/res_ph3_",drug_family,"_RCT_TS",TS,".RData"))
  out <- purrr::map_dfr(c("Main",1:8),function(sce)
{
  load(paste0("results/res_EffTox/res_ph1_",
              drug_family,
              "_EffTox_sce",
              sce,".RData"))
    res_ph1_obj <- res_ph1_EffTox
    res <- combine_res(res_ph1 = res_ph1_obj,
                       res_ph2 = res_ph2_simon2stage,
                       res_ph3 = res_ph3_RCT)
    out <- data.frame(
      Ph123_PoS = res$`Phase I-III probability of significance`,
      Ph23_PoTS = res$`Phase II&III probability of toxicity suspension`
    ) |>
      mutate(Scenario = sce) |>
      mutate(Scenario = factor(Scenario,levels = c("Main",1:8)))
    return(out)
  })
  out <- out |>
    mutate( TS_rule = TS
    )
  return(out)
})
res_ph123_df <- res_ph123_df |>
  mutate(TS_rule = case_when(
    is.na(TS_rule) ~ "None",
    TRUE ~ paste0(TS_rule*100,"% DLT")
  ))

drug_family_label <- case_when(
      drug_family == "immuno_fixed_OBD3_MTD5" ~ "immuno-like profile OBD=3",
      drug_family == "immuno_fixed_OBD4_MTD5" ~ "immuno-like profile OBD=4",
      drug_family == "chemo_fixed_MTD5" ~ "chemo-like profile MTD=5"
    )
p <- res_ph123_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph123_PoS,color=factor(Scenario),
             group = factor(Scenario)))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Phase I-III Probability of Significance (PoS)",
       title = drug_family_label,
       color = "Contour setting")+
  # scale_color_viridis_d()+
  scale_color_manual(values = c( "red","#440154FF", "#46337EFF", "#365C8DFF",
                                "#277F8EFF", "#1FA187FF", "#4AC16DFF" ,"#9FDA3AFF", "#FDE725FF"))+
  theme(
    legend.position = "right"
  )
  return(p)
},simplify = FALSE)
# immuno-like
ggarrange(plotlist = p_list[1:2],ncol = 2,
          common.legend = TRUE)
# chemo-like
p_list[[3]]
```
