---
title: "Comparison of Phase I dose-finding methods"
author: "Yongxi Long"
format: html
  self-contained: TRUE
code-fold: true
editor: visual
---

## Aim

To evaluate the impact of phase I dose-finding methods on later phase success rate. We compare the following 3 dose-finding method

-   3+3: rule-based method
-   BOIN: model-assisted method, with some toxicity control
-   CRM: model-based method, with some toxicity control

We'll look at two types of drug profiles, chemotherapy-like profiles vs. immunotherapy-like profiles

```{r,warning=FALSE,message=FALSE}
rm(list = ls())
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsci)
  library(RColorBrewer)
  library(trialr)
  library(clinfun)
  library(ggExtra)
  library(ggridges)
  library(ggpubr)
})
source("functions/combine_res.R")
```

# Load data

```{r,eval=FALSE}
drug_families <- c("immuno_early_plateau", "immuno_late_plateau",
                   "chemo_flat_efficacy", "chemo_steep_efficacy")
drug_family <- drug_families[1]
load(paste0("data/simulated_data/",drug_family,".RData"))

# function for making drug family name label
make_drug_label <- function(drug_family)
{
  drug_label <- case_when(
    drug_family == "immuno_early_plateau" ~ "Immuno-like profile, early plateau",
    drug_family == "immuno_late_plateau" ~ "Immuno-like profile, late plateau",
    drug_family == "chemo_flat_efficacy" ~ "Chemo-like profile, flat efficacy",
    drug_family == "chemo_steep_efficacy" ~ "Chemo-like profile, steep efficacy",
  )
  return(drug_label)
}
```

## Phase I

```{r,eval=FALSE}
# trial design parameters
# sample size
N <- 30
# target DLT
target_DLT <- 0.3
```

### 3+3 design

```{r,eval=FALSE}
source("functions/threep3.R")
res_ph1_threep3 <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
res <- run_threep3(data = drug_data[[i]],
                   N=30)
res_ph1_threep3[[i]] <- res
}
save(res_ph1_threep3,
     file = paste0("results/res_ph1_threep3.RData"))
combine_res(res_ph1 = res_ph1_threep3)
mean(sapply(res_ph1_threep3, function(res) sum(res$summary_tab$num_pts)))
```

### BOIN design

```{r,eval=FALSE}
source("functions/BOIN.R")
# cohort size
cohort_size= 3
# BOIN design
res_ph1_BOIN <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
res <- run_BOIN(data = drug_data[[i]],
                 N=N,
                 cohort_size = cohort_size,
                 target_DLT=target_DLT,
                 tox_bound = 0.95)
res_ph1_BOIN[[i]] <- res
}
save(res_ph1_BOIN,
     file = paste0("results/res_ph1_BOIN.RData"))
combine_res(res_ph1 = res_ph1_BOIN)
```

### CRM

```{r,eval=FALSE}
source("functions/CRM.R")
# cohort size
cohort_size= 1
# toxicity boundary
tox_bound <- 0.9
# from paper Chu et. al
# Unifying CRM and EWOC designs for Phase I cancer clinical trials
# CRM without overdose control
res_ph1_CRM <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  res <- run_CRM(data=drug_data[[i]],
                    N=N,
                    cohort_size = cohort_size,
                    target_DLT=target_DLT,
                    tox_bound = tox_bound,
                    model="empiric")
  res_ph1_CRM[[i]] <- res
  print(i)
}
save(res_ph1_CRM,file = paste0("results/res_ph1_CRM.RData"))
combine_res(res_ph1 = res_ph1_CRM)
```

### EffTox design

```{r,eval=FALSE}
source("functions/EffTox.R")
# trial design parameters
# sample size
N <- 30
# cohort size
cohort_size <- 3
# Threshold for selecting desirable doses
efficacy_hurdle <- 0.5 # desired response rate
toxicity_hurdle <- 0.3 # target DLT
p_e <- p_t <- 0.05
# the Contour
eff0 <- 0.4 # smallest rsp acceptable if there is no toxicity, set to slightly smaller than SOC, p0 = 0.4
tox1 <- 0.7 # highest DLT acceptable if the rsp is 100%, set to 150% of the target DLT
# middle neutral utility value
# slight more toxic than target, and slight less efficacious than target
eff_star <- 0.65
tox_star <- 0.25
p_EffTox <-  trialr::efftox_solve_p(eff0,tox1,eff_star,tox_star)
# prior guess of DLT/RSP prob for each dose
pi_T <- dfcrm::getprior(halfwidth = 0.05,target = 0.3,nu=3,nlevel=6,model="empiric")
pi_E <- dfcrm::getprior(halfwidth = 0.05,target = 0.5,nu=3,nlevel=6,model="empiric")
# effective sample size for the prior
ess_T <- ess_E <- 1
# get normal prior hyperparameters for the EffTox model
# it can take some time for iterative calculation
# the dose levels are already scaled between 0-1, it does 
# not impact the prior hyperparamters that much
# so for saving time, will use the same (average) actual dose
# levels for all simulation iteration, although the actual dose
# levels are different for each iteration
priors_EffTox <- get_efftox_priors(doses = c(0.1000000, 
                                             0.2000000, 
                                             0.3300000, 
                                             0.5016000, 
                                             0.7022400, 
                                             0.9339792),
                                   pi_T = pi_T,
                                   ess_T = ess_T,
                                   pi_E = pi_E,
                                   ess_E = ess_E)
```

```{r,eval=FALSE}
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
res_ph1_EffTox <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  dat <- drug_data[[i]]
  res <- suppressWarnings({
    run_EffTox(
      data = dat,
      N=N,
      cohort_size=cohort_size,
      efficacy_hurdle = efficacy_hurdle, 
      toxicity_hurdle = toxicity_hurdle,
      p_e = p_e, p_t = p_t,
      eff0 = eff0, tox1 = tox1,
      eff_star = eff_star, tox_star = tox_star,
      priors = priors_EffTox)
  })
  res_ph1_EffTox[[i]] <- res
  print(i)
}
save(res_ph1_EffTox,file = paste0("results/res_ph1_EffTox_",drug_family,".RData"))
}
combine_res(res_ph1 = res_ph1_EffTox)
```

### BOIN12 design

```{r,eval=FALSE}
source("functions/BOIN12.R")
# sample size
N <- 30
# the toxicity upper limit
phi_t <- 0.35
# the efficacy lower limit
phi_e <- 0.5
# utility for outcome no tox no response
u2 <- 40
# utility for outcome tox yes and response yes
u3 <- 60
# when tox is within bounds, 
# stop exploring higher doses when n at dose is greater than or equal to this value.
n_star1 <- 6
# dose-exploration rule, go to the next higher dose if more than this number of patients have been treated at current dose and the current dose is safe enough
# and the next higher dose has not yet been explored
n_star2 <- 8
```

```{r,eval=FALSE}
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
res_ph1_BOIN12 <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  dat <- drug_data[[i]]
  res <- suppressWarnings({
    run_BOIN12(data = dat,
               N= N,
               cohort_size = 3,
               phi_t = phi_t, phi_e = phi_e,
               u2 = u2, u3 = u3,
               n_star1 = n_star1, n_star2 = n_star2)
  })
  res_ph1_BOIN12[[i]] <- res
}
save(res_ph1_BOIN12,file = paste0("results/res_ph1_BOIN12_",drug_family,".RData"))
}
combine_res(res_ph1 = res_ph1_BOIN12)
```

### Evaluate Phase I dose selection

```{r,echo=FALSE}
drug_families <- c("immuno_early_plateau", "immuno_late_plateau",
                   "chemo_flat_efficacy", "chemo_steep_efficacy")
designs <- c("threep3", "BOIN", "CRM")
RP2D_summary_MTD <- purrr::map_dfr(designs, function(design) {
  # Load the results
  load(paste0("results/res_ph1_", design, ".RData"))
  # Retrieve the actual object (e.g., res_ph1_threep3)
  res_ph1_obj <- get(paste0("res_ph1_", design))
  res <- combine_res(res_ph1 = res_ph1_obj)
  res <- res$`Table of RP2D` |>
    select(dose_level,prob_selected) |>
    mutate(Design = design) |>
    mutate(Design = factor(Design,levels=c("threep3","BOIN","CRM","EffTox","BOIN12"),
                           labels = c("Threep3","BOIN","CRM","EffTox","BOIN12"))) |>
    mutate(Drug_family = "fixed")
  return(res)
})
designs <- c("BOIN12","EffTox")
RP2D_summary_OBD <- purrr::map_dfr(designs,function(design)
  {
  df <- purrr::map_dfr(drug_families,function(drug_family)
{
  load(paste0("results/res_ph1_",design,"_",drug_family,".RData"))
  # Retrieve the actual object (e.g., res_ph1_threep3)
  res_ph1_obj <- get(paste0("res_ph1_", design))
  res <- combine_res(res_ph1 = res_ph1_obj)
  res <- res$`Table of RP2D` |>
    select(dose_level,prob_selected) |>
    mutate(Design = design) |>
    mutate(Design = factor(Design,levels=c("threep3","BOIN","CRM","EffTox","BOIN12"),
                           labels = c("Threep3","BOIN","CRM","EffTox","BOIN12"))) |>
    mutate(Drug_family = drug_family)
  return(res)
})
  return(df)
})
p1 <- RP2D_summary_MTD |>
  ggplot(aes(x = Design, y = prob_selected, fill = factor(dose_level))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    x = "Design",
    y = "Probability of Selection",
    fill = "RP2D",
    title = "Univariate Designs"
  )+
  geom_text(aes(label = round(prob_selected, 2)),  # text labels
            position = position_dodge(width = 0.8), 
            vjust = -0.3,                          # vertical position (above bar)
            size = 3.5,
            angle = 45)+
  scale_fill_viridis_d(option = "viridis",na.value="grey80")+
  theme(
    title = element_text(size=8),
    axis.text.x = element_text(size=7),
    axis.title = element_text(size=8)
  )+
  ylim(0,0.8)
p1
p2 <- RP2D_summary_OBD |>
  filter(Design=="EffTox") |>
  ggplot(aes(x = Drug_family, y = prob_selected, fill = factor(dose_level))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    x = "Drug family",
    y = "Probability of Selection",
    fill = "RP2D",
    title = "EffTox"
  )+
  geom_text(aes(label = round(prob_selected, 2)),  # text labels
            position = position_dodge(width = 0.8), 
            vjust = -0.3,                          # vertical position (above bar)
            size = 3,
            angle = 45)+
  scale_fill_viridis_d(option = "viridis",na.value="grey80")+
  theme(
    title = element_text(size=8),
    axis.text.x = element_text(size=8,angle = 10),
    axis.title = element_text(size=8)
  )+
  ylim(0,0.7)

p3 <- RP2D_summary_OBD |>
  filter(Design=="BOIN12") |>
  ggplot(aes(x = Drug_family, y = prob_selected, fill = factor(dose_level))) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    x = "Drug family",
    y = "Probability of Selection",
    fill = "RP2D",
    title = "BOIN12"
  )+
  geom_text(aes(label = round(prob_selected, 2)),  # text labels
            position = position_dodge(width = 0.8), 
            vjust = -0.3,                          # vertical position (above bar)
            size = 3,
            angle = 45)+
  scale_fill_viridis_d(option = "viridis",na.value="grey80")+
  theme(
    title = element_text(size=8),
     axis.text.x = element_text(size=8,angle = 10),
    axis.title = element_text(size=8)
  )+
  ylim(0,0.7)
ggarrange(p2,p3,ncol=2,common.legend = TRUE,
          legend = "top")
# ggsave(filename = paste0("C:\\Users\\ylong\\OneDrive - LUMC\\Projects\\Impacts_early_phase_trial_methods\\plots\\Phase1_dose_selection.png"),
#        plot = plot,
#        dpi=320,
#        width = 7,
#        height = 5,
#        units = "in")
```

## Phase II

### Simon's two-stage design

We first supply clinical information to get design parameters for the Simon's two-stage design

```{r,eval=FALSE}
source("functions/simon2stage.R")
# previous too large Phase II design under p0=0.4, p1=0.5, alpha=0.05
# design_para <- simon2stage_design(p0=0.4,
#                                   p1=0.5,
#                                   alpha = 0.05)
# common Phase II trials have n <80
design_para <- simon2stage_design(p0=0.4,
                                  p1=0.5,
                                  beta = 0.4,
                                  alpha = 0.1)
```

```{r,eval=FALSE}
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
# different extent of toxicity control at Phase II
# noTS: no toxicity induced trial suspension
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
for(TS in TS_rules)
{
  res_ph2_simon2stage <- vector(mode = "list",length = length(drug_data))
  for (i in 1:length(drug_data))
  {
    res <- run_simon2stage(
    data = drug_data[[i]],
    design_para = design_para,
    suspend_threshold = TS,
    suspend_window = 20)
    res_ph2_simon2stage[[i]] <- res
  }
save(res_ph2_simon2stage,file = paste0("results/res_ph2_s2s_TS",TS,"_",drug_family,".RData"))
}
}
```

### Combine Phase I-II results

```{r,eval=FALSE}
drug_families <- c("immuno_early_plateau", "immuno_late_plateau",
                   "chemo_flat_efficacy", "chemo_steep_efficacy")
plot_list_ph12 <- sapply(drug_families, function(drug_family)
  {
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
designs <- c("threep3", "BOIN", "CRM","EffTox","BOIN12")
# for CRM design
# different extend of toxicity control in Phase II
res_ph12_df <- purrr::map_dfr(TS_rules,function(TS)
  {
  load(paste0("results/res_ph2_s2s_TS",TS,"_",drug_family,".RData"))
  
  out <- purrr::map_dfr(designs,function(design)
  {
    if(design %in% c("EffTox","BOIN12"))
    {
      load(paste0("results/res_ph1_",design,"_",drug_family,".RData"))
    } else
    {
      load(paste0("results/res_ph1_", design, ".RData"))
    }
    res_ph1_obj <- get(paste0("res_ph1_", design))
    res <- combine_res(res_ph1 = res_ph1_obj,res_ph2 = res_ph2_simon2stage)
    out <- data.frame(
      Ph12_PoS = res$`Phase I-II probability of significance`,
      Ph2_PoTS = res$`Phase II probability of toxicity suspension`
    ) |>
      mutate(Design = design)
    return(out)
  })
  out <- out |>
    mutate( TS_rule = TS
    )
  return(out)
})
res_ph12_df <- res_ph12_df |>
  mutate(TS_rule = case_when(
    is.na(TS_rule) ~ "None",
    TRUE ~ paste0(TS_rule*100,"% DLT")
  ))

p1 <- res_ph12_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph12_PoS,color=Design,
             group = Design))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Phase I-II Probability of Significance (PoS)",
     #  title = "Phase I-II results",
       color = "Dose-finding method")+
  scale_color_nejm()

p2 <- res_ph12_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph2_PoTS,color=Design,
             group = Design))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Phase II Probability of Toxicity Suspension (PoTS)",
    #   title = "Phase I-II results",
       color = "Dose-finding method")+
  scale_color_nejm()

p12 <- ggarrange(p1,p2,ncol = 2,legend = "top",
          common.legend = TRUE)

plot <- annotate_figure(
  p = p12,
  top = text_grob(paste0("Drug profile: ",drug_family))
)
return(plot) 
},simplify = FALSE)
plot_list_ph12[[4]]
```

## Phase III

### RCT design

We use the RCT_design function to help us do a sample size calculation

```{r,eval=FALSE}
source("functions/RCT.R")
design_para <- RCT_design(hazard0 = log(2)/14,hazard1 = log(2)/19,
                          t_FU = 60, t_accrual = 12)
```

```{r,eval=FALSE}
for(drug_family in drug_families)
{
load(paste0("data/simulated_data/",drug_family,".RData"))
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
for(TS in TS_rules)
{
res_ph3_RCT <- vector(mode = "list",length = length(drug_data))
for (i in 1:length(drug_data))
{
  res <- run_RCT(
  data = drug_data[[i]],
  design_para = design_para,
  suspend_threshold = TS,
  suspend_window = 30)
  res_ph3_RCT[[i]] <- res
  print(i)
}
save(res_ph3_RCT,file = paste0("results/res_ph3_RCT_TS",TS,"_",drug_family,".RData"))
}
}
```

### Combine Phase I-III results

```{r}
drug_families <- c("immuno_early_plateau", "immuno_late_plateau",
                   "chemo_flat_efficacy", "chemo_steep_efficacy")
TS_rules <- list(NULL,0.45,0.40,0.35,0.30)
designs <- c("threep3", "BOIN", "CRM","EffTox","BOIN12")
plot_list_ph123 <- sapply(drug_families, function(drug_family)
  {
res_ph123_df <- purrr::map_dfr(TS_rules,function(TS)
  {
  load(paste0("results/res_ph2_s2s_TS",TS,"_",drug_family,".RData"))
  load(paste0("results/res_ph3_RCT_TS",TS,"_",drug_family,".RData"))
  out <- purrr::map_dfr(designs,function(design)
  {
    if(design %in% c("EffTox","BOIN12"))
    {
      load(paste0("results/res_ph1_",design,"_",drug_family,".RData"))
    } else
    {
      load(paste0("results/res_ph1_", design, ".RData"))
    }
    res_ph1_obj <- get(paste0("res_ph1_", design))
    res <- combine_res(res_ph1 = res_ph1_obj,
                       res_ph2 = res_ph2_simon2stage,
                       res_ph3 = res_ph3_RCT)
    out <- data.frame(
      Ph123_PoS = res$`Phase I-III probability of significance`,
      Ph23_PoTS = res$`Phase II&III probability of toxicity suspension`
    ) |>
      mutate(Design = design)
    return(out)
  })
  out <- out |>
    mutate( TS_rule = TS
    )
  return(out)
})
res_ph123_df <- res_ph123_df |>
  mutate(TS_rule = case_when(
    is.na(TS_rule) ~ "None",
    TRUE ~ paste0(TS_rule*100,"% DLT")
  ))

p1 <- res_ph123_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph123_PoS,color=Design,
             group = Design))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Probability of Success (PoS)",
      title = make_drug_label(drug_family),
       color = "")+
  scale_color_nejm()+
  theme(
     title = element_text(size = 9),
     axis.title = element_text(size=9),
     axis.text = element_text(size=8)
  )+
  ylim(0,0.6)

p2 <- res_ph123_df |>
  mutate(TS_rule_f = factor(TS_rule,
                            levels=c("None","45% DLT","40% DLT","35% DLT","30% DLT"))) |>
  ggplot(aes(x=TS_rule_f,y=Ph23_PoTS,color=Design,
             group = Design))+
  geom_point()+
  geom_line()+
  theme_bw()+
  labs(x="Toxicity suspension threshold",
       y="Probability of Toxicity Suspension (PoTS)",
       title = make_drug_label(drug_family),
       color = "")+
  scale_color_nejm()+
  theme(
     title = element_text(size = 9),
     axis.title.x = element_text(size=9),
     axis.title.y = element_text(size=8),
     axis.text = element_text(size=8)
  )+
  ylim(0,0.7)
# p12 <- ggarrange(p1,p2,ncol = 2,legend = "top",
#           common.legend = TRUE)
# 
# plot <- annotate_figure(
#   p = p12,
#   top = text_grob(paste0("Drug profile: ",drug_family))
# )
# return(plot)
return(list("PoS"=p1,"PoTS"=p2))
},simplify = FALSE)
plot_list_ph123_PoS <- sapply(plot_list_ph123, function(i) i[["PoS"]],simplify=FALSE)
plot_list_ph123_PoTS <- sapply(plot_list_ph123, function(i) i[["PoTS"]],simplify=FALSE)
plot_PoS <- ggarrange(plotlist = plot_list_ph123_PoS,ncol=2,nrow=2,common.legend = TRUE,
          legend = "right")
plot_PoTS <- ggarrange(plotlist = plot_list_ph123_PoTS,ncol=2,nrow=2,common.legend = TRUE,
          legend = "right")
ggsave(filename = paste0("C:\\Users\\ylong\\OneDrive - LUMC\\Projects\\Impacts_early_phase_trial_methods\\plots\\plot_PoTS.png"),
       plot = plot_PoTS,
       dpi=320,
       width = 7,
       height = 5,
       units = "in")
```
