---
title: "Comparison of Phase I/II Design Strategies"
author: "Yongxi Long"
format: html
  self-contained: TRUE
editor: visual
---

## Aim

To evaluate the impact of early phase design strategies on later phase success rate. We mainly compare two strategies to evaluate whether it is more efficient to invest more in Phase I or in Phase II:

-   Consider both toxicity and efficacy in Phase I (univariate CRM), carry one dose into Phase II
-   Consider only toxicity in Phase I (bivariate CRM), carry two doses into Phase II

We'll look at three types of drug profiles, where the dose-efficacy/survival relationships plateau before/around/after the MTD.

```{r,warning=FALSE,message=FALSE}
rm(list = ls())
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggsci)
  library(RColorBrewer)
  library(trialr)
  library(clinfun)
  library(ggExtra)
  library(ggridges)
})
```

## Load data

```{r}
drug_family <- "immuno"
load(paste0("data/simulated_data_",drug_family,".RData"))
```

## Phase I

### CRM design

```{r}
source("functions/CRM.R")
# trial design parameters
# sample size
N <- 30
# cohort size
cohort_size= 1
# target DLT
target_DLT <- 0.3
# overdose control
# toxicity bound, if Pr(toxcitiy prob at dose j > target DLT) > tox_bound
# or Equivalently Pr(toxcitiy prob at dose j < target DLT) < 1 - tox_bound
# this dose will be inadmissible even if it may be recommended by the model
tox_bound <- 0.25
```

```{r}
# CRM without overdose control
res_ph1_CRM <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  res <- run_CRM(data=drug_data[[i]],
                    N=N,
                    cohort_size = cohort_size,
                    target_DLT=target_DLT,
                    tox_bound = tox_bound,
                    model="empiric")
  res_ph1_CRM[[i]] <- res
  print(i)
}
save(res_ph1_CRM,file = paste0("results/res_ph1_",drug_family,"_CRM_OC25.RData"))
```

### EffTox design

We first need to setup the design parameters for EffTox

```{r}
source("functions/EffTox.R")
# trial design paramters
# sample size
N <- 30
# cohort size
cohort_size <- 3
# Threshold for selecting desirable doses
efficacy_hurdle <- 0.3
toxicity_hurdle <- 0.6
p_e <- p_t <- 0.1
# the Contour
eff0 <- 0.35 # smallest rsp acceptable if there is no toxicity, set to same as SOC
tox1 <- 0.45 # highest DLT acceptable if the rsp is 100%, set to 150% of the target DLT
eff_star <- 0.4
tox_star <- 0.25
p_EffTox <- efftox_solve_p(eff0,tox1,eff_star,tox_star)
# prior guess of DLT/RSP prob for each dose
pi_T <- dfcrm::getprior(halfwidth = 0.05,target = 0.3,nu=3,nlevel=5,model="empiric")
pi_E <- dfcrm::getprior(halfwidth = 0.05,target = 0.45,nu=3,nlevel=5,model="empiric")
# effective sample size for the prior
ess_T <- ess_E <- 0.5
# get normal prior hyperparameters for the EffTox model
# it can take some time for iterative calculation
# the dose levels are already scaled between 0-1, it does 
# not impact the prior hyperparamters that much
# so for saving time, will use the same (average) actual dose
# levels for all simulation iteration, although the actual dose
# levels are different for each iteration
priors_EffTox <- get_efftox_priors(doses = c(0.12,0.25,0.41,0.63,0.87),
                                   pi_T = pi_T,
                                   ess_T = ess_T,
                                   pi_E = pi_E,
                                   ess_E = ess_E)
```

Perform the EffTox design on the 1000 drug profiles

```{r}
res_ph1_EffTox <- vector(mode = "list",length = length(drug_data))
for(i in 1:length(drug_data))
{
  dat <- drug_data[[i]]
  res <- run_EffTox(
      data = dat,
      N=N,
      cohort_size=cohort_size,
      efficacy_hurdle = efficacy_hurdle, toxicity_hurdle = toxicity_hurdle,
      p_e = p_e, p_t = p_t,
      eff0 = eff0, tox1 = tox1,
      eff_star = eff_star, tox_star = tox_star,
      priors = priors_EffTox)
  res_ph1_EffTox[[i]] <- res
  print(i)
}
save(res_ph1_EffTox,file = paste0("results/res_ph1_",drug_family,"_EffTox.RData"))
```

### Examine Phase I results

#### Summary table of dose selection

```{r}
RP2D_CRM <- sapply(res_ph1_CRM,function(x) x$RP2D)
RP2D_EffTox <- sapply(res_ph1_EffTox, function(x) x$RP2D)
RP2D_summary <- data.frame(RP2D = factor(c(RP2D_CRM,RP2D_EffTox),levels=1:5),
                           Design = rep(c("CRM","EffTox"),
                                        c(length(RP2D_CRM),
                                          length(RP2D_EffTox)))) 
RP2D_summary |>
  group_by(Design, RP2D,.drop=FALSE) |>
  summarize(n=n()) |>
  mutate(Prob_selected = n/sum(n))|>
  select(-n) |>
  ungroup()
```


#### Plot of the true OBD vs. selected by CRM/EffTox

```{r,warning=FALSE}
nsim <- length(drug_data)
# see which dose has been selected on the true tox - eff panel
df_CRM_selected <- df_EffTox_selected <- df_OBD_true <- data.frame(
  iteration = 1:nsim,
  dose = rep(NA,nsim),
  prob_DLT = rep(NA,nsim),
  prob_RSP = rep(NA,nsim)
)
for(i in 1:nsim)
{
  # determine the true OBD
  # get the utility of efficacy & toxicity probability pairs
  rsp_rate <- drug_data[[i]]$profile_info$rsp_rate
  DLT_rate <- drug_data[[i]]$profile_info$DLT_rate
  utility <- efftox_utility(p=p_EffTox,
                            eff0=eff0,
                            tox1=tox1,
                            prob_eff = rsp_rate,
                            prob_tox = DLT_rate)
  OBD_true <- which.max(utility)
  df_OBD_true[i,2:4] <- c(OBD_true,
                          DLT_rate[OBD_true],
                          rsp_rate[OBD_true])
  
  dose_EffTox <- res_ph1_EffTox[[i]]$RP2D
  df_EffTox_selected[i,2:4] <- c(dose_EffTox,
                                 ifelse(is.na(dose_EffTox),NA,DLT_rate[dose_EffTox]),
                                 ifelse(is.na(dose_EffTox),NA,rsp_rate[dose_EffTox]))
  dose_CRM <- res_ph1_CRM[[i]]$RP2D
  df_CRM_selected[i,2:4] <- c(dose_CRM,
                              ifelse(is.na(dose_CRM),NA,DLT_rate[dose_CRM]),
                              ifelse(is.na(dose_CRM),NA,rsp_rate[dose_CRM]))
}
df_OBD_true$type <- "True OBD"
df_EffTox_selected$type <- "EffTox selected"
df_CRM_selected$type <- "CRM selected"
df_comb <- rbind(df_OBD_true,df_EffTox_selected,df_CRM_selected)
# percentage of selecting the OBD correctly
# CRM
prob_OBD_CRM <- mean(df_OBD_true$dose == df_CRM_selected$dose,na.rm = TRUE)
# EffTox
prob_OBD_EffTox <- mean(df_OBD_true$dose == df_EffTox_selected$dose,na.rm = TRUE)
p1 <- df_comb |>
  ggplot(aes(x=prob_RSP,y=prob_DLT,color=type))+
  geom_point(alpha=0.3)+
  theme_bw()+
  geom_vline(xintercept = 0.45,linetype=2,color="seagreen")+
  geom_hline(yintercept = 0.3,linetype=2,color="indianred3")+
  ylim(c(0,1))+
  xlim(c(0.3,0.6))+
  labs(x="Probability of response",y="Probability of DLT")+
  annotate("text",label=paste0("Prob of selecting OBD by EffTox = ",round(prob_OBD_EffTox,2)),x=0.5, y =0.95)+
  annotate("text",label=paste0("Prob of selecting OBD by CRM = ",round(prob_OBD_CRM,2)),x=0.5, y =0.9)
ggMarginal(p1,
           type = "density",
           groupFill = TRUE)
```

#### Risk/efficacy exposure assessment

See the true DLT rates/response rates of doses that patients were assigned to 

```{r}
DLT_rate_all <- RSP_rate_all <- design_all <- numeric()

for(i in 1:length(drug_data)) {
  
  # CRM
  dose_history_CRM <- res_ph1_CRM[[i]]$dose_history
  DLT_rate_vec <- drug_data[[i]]$profile_info$DLT_rate[dose_history_CRM]
  RSP_rate_vec <- drug_data[[i]]$profile_info$rsp_rate[dose_history_CRM]
  
  DLT_rate_all <- c(DLT_rate_all, DLT_rate_vec)
  RSP_rate_all <- c(RSP_rate_all, RSP_rate_vec)
  design_all <- c(design_all, rep("CRM", length(DLT_rate_vec)))
  
  # EffTox
  dose_history_EffTox <- res_ph1_EffTox[[i]]$dose_history
  DLT_rate_vec <- drug_data[[i]]$profile_info$DLT_rate[dose_history_EffTox]
  RSP_rate_vec <- drug_data[[i]]$profile_info$rsp_rate[dose_history_EffTox]
  
  DLT_rate_all <- c(DLT_rate_all, DLT_rate_vec)
  RSP_rate_all <- c(RSP_rate_all, RSP_rate_vec)
  design_all <- c(design_all, rep("EffTox", length(DLT_rate_vec)))
}

# Construct final data frame
patient_dosing_history <- data.frame(
  DLT_rate = DLT_rate_all,
  rsp_rate = RSP_rate_all,
  design = design_all
)

patient_dosing_summary <- patient_dosing_history |>
  group_by(design) |>
  summarise(mean_DLT = mean(DLT_rate),
            prob_overdosing = mean(DLT_rate > 0.3),
            mean_rsp = mean(rsp_rate),
            prob_futile = mean(rsp_rate < 0.45))
# DLT rates
p1 <- patient_dosing_history |>
ggplot(aes(x = DLT_rate, y = design)) +
  geom_density_ridges() +
  geom_vline(xintercept = 0.30, linetype = "dashed") +
  labs(x = "DLT rate of doses assigned to patients", y = "Design")+
  theme_bw()+
   annotate("text",x=0.6,y=3.5,label=paste0("EffTox: P(dosing a patient at > target DLT)=",round(patient_dosing_summary$prob_overdosing[2],2)))+
  annotate("text",x=0.6,y=3.2,label=paste0("CRM: P(dosing a patient at > target DLT)=",round(patient_dosing_summary$prob_overdosing[1],2)))

# Response rates
p2 <- patient_dosing_history |>
ggplot(aes(x = rsp_rate, y = design)) +
  geom_density_ridges() +
  geom_vline(xintercept = 0.45, linetype = "dashed") +
  labs(x = "Response rate of doses assigned to patients", y = "Design")+
  theme_bw()+
   annotate("text",x=0.2,y=3.2,label=paste0("EffTox: P(dosing a patient at < target response)=",round(patient_dosing_summary$prob_futile[2],2)))+
  annotate("text",x=0.2,y=2.9,label=paste0("CRM: P(dosing a patient at < target response)=",round(patient_dosing_summary$prob_futile[1],2)))

ggpubr::ggarrange(p1,p2,ncol=1,nrow=2)
```

## Phase II

### Simon's two-stage design

We first supply clinical information to get design parameters for the Simon's two-stage design

```{r}
design_para <- simon2stage_design(p0=0.35,
                                  p1=0.45)
```

```{r}
res_ph2_simon2stage <- vector(mode = "list",length = length(drug_data))
for (i in 1:length(drug_data))
{
  res <- run_simon2stage(
  data = drug_data[[i]],
  design_para = design_para)
  
  res_ph2_simon2stage[[i]] <- res
}
```

## Combine Phase I-II results

```{r}

```
