---
title: "Generate fixed drug profiles for simulation"
author: "Yongxi Long"
format: 
  html:
    self-contained: true
editor: visual
---

## Aim

This is to generate drug profiles of different family:

-   Chemotherapy-like profiles: the toxicity curves and efficacy curves parallel
-   Immunotherapy-lie profiles: the efficacy curves plateau after some dose while the toxicity curves are monotonnically increasing.

Here we generate data from a fixed profile, i.e., all iterations of the simulation evaluate the same drug profile

```{r,warning=FALSE,echo=FALSE}
rm(list = ls())
suppressPackageStartupMessages(
  {
    library(ggplot2)
    library(dplyr)
    library(ggpubr)
    library(simsurv)
  }
)
source("data_generating_functions.R")
```

## Function skeletons

1.  For toxicity, we use more a monotonic function, the scaled logistic:

$$P(Y=1\mid d) = L \frac{1-e^{-kd}}{1-e^{-k}}$$

-   $k$ adjusts the curve growth rate.

2.  For response and survival, we use an S-shaped skeleton to obtain a plateau:

$$P(Y=1\mid d) = L \frac{d^a}{d^a+(1-d)^b}$$

a and b are shape control parameters:

-   if a=b, the S-curve is symmetric around d = 0.5

-   if $a<b$, the S-curve rises faster initially, otherwise the curve starts very flat and rises more at the end.

-   For dose-survival, we first obtain some value between 0 and 1 from the above function, then scale them by some constant to have proper units (e.g., days, months, years). We also add an offset to make sure that the survival time at dose zero is a positive value (otherwise it will be zero).

## Parameter setup for four drug profiles

```{r,echo=TRUE}
# dose selection
ndose <- 6
lowest_dose <- 0.1
dose_levels <- lowest_dose*c(1,2,2*1.65,2*1.65*1.52,2*1.65*1.52*1.4,2*1.65*1.52*1.4*1.33)
profile_parameter_list <- list(
  "immuno_early_plateau" = list(k_tox = -3.5,L_tox = 0.8,
                                L_rsp=0.5, L_surv = 0.6,
                                a_rsp = 1, b_rsp = 12,
                                a_surv = 0.1, b_surv = 12,
                                scale_surv = 20, offset_surv = 8),
  "immuno_late_plateau" =  list(k_tox = -3.5,L_tox = 0.8,
                                L_rsp=0.5, L_surv = 0.6,
                                a_rsp = 2, b_rsp = 8,
                                a_surv = 2, b_surv = 8,
                                scale_surv = 20, offset_surv = 8),
  "chemo_flat_efficacy" = list(k_tox = -3.5, L_tox = 0.8,
                               L_rsp = 0.58, L_surv = 0.68,
                               k_rsp = 2, k_surv = 2,
                               scale_surv = 20, offset_surv = 8),
  "chemo_steep_efficacy" = list(k_tox = -3.5, L_tox = 0.8,
                               L_rsp = 0.85, L_surv = 1,
                               k_rsp = -1, k_surv = -1,
                               scale_surv = 20, offset_surv = 8)
)
```

## Immuno-like drug profiles

In this type of drug profiles (immuno/targeted therapies), the efficacy curves plateau before MTD. Namely, the optimal biological dose (OBD) is smaller than the maximum tolerable dose (@fig-dose-tox-rsp-surv). We set $a<b$ for dose-response, dose-survival curves to achieve this. For dose-(median) survival curve, we add scale and offset parameter in the function to make the survival time in proper range and has a minimum positive value.

How earlier the efficacy plateau is relative to the MTD is a tuning parameter. We'll evaluate the case where OBD is much separated from MTD (@fig-immuno-OBD3-MTD5), as well as where the OBD and the MTD is close (@fig-immuno-OBD4-MTD5).

```{r,echo=FALSE}
#| label: fig-immuno-OBD3-MTD5
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a immunotherapy-like profile wtih an early-plateau of efficacy, the OBD is dose 3, which is very separated from MTD which is dose 5."
# see the drug profile
drug_family <- "immuno_early_plateau"
profile_parameter <- profile_parameter_list[[drug_family]]
plot_immuno_OBD3 <- plot_drug_profiles(dose_levels = dose_levels,
                   k_tox=profile_parameter$k_tox,L_tox=profile_parameter$L_tox,
                   a_rsp=profile_parameter$a_rsp,b_rsp=profile_parameter$b_rsp,
                   L_rsp=profile_parameter$L_rsp,
                   a_surv=profile_parameter$a_surv,b_surv=profile_parameter$b_surv,
                   L_surv=profile_parameter$L_surv,
                   scale_surv=profile_parameter$scale_surv,
                   offset_surv=profile_parameter$offset_surv,
                   plot_title = "Immuno-like profile, early plateau")
plot_immuno_OBD3
```

```{r,echo=FALSE}
#| label: fig-immuno-OBD4-MTD5
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a immunotherapy-like profile with a late plateau of efficacy, the OBD is dose 4, which is close to MTD which is dose 5."
# see the drug profile
drug_family <- "immuno_late_plateau"
profile_parameter <- profile_parameter_list[[drug_family]]
plot_immuno_OBD4 <- plot_drug_profiles(dose_levels = dose_levels,
                   k_tox=profile_parameter$k_tox,L_tox=profile_parameter$L_tox,
                   a_rsp=profile_parameter$a_rsp,b_rsp=profile_parameter$b_rsp,
                   L_rsp=profile_parameter$L_rsp,
                   a_surv=profile_parameter$a_surv,b_surv=profile_parameter$b_surv,
                   L_surv=profile_parameter$L_surv,
                   scale_surv=profile_parameter$scale_surv,
                   offset_surv=profile_parameter$offset_surv,
                   plot_title = "Immuno-like profile, late plateau")
plot_immuno_OBD4
```

## Chemo-like drug profiles

In this type of drug profiles (chemo therapies), the efficacy curves parallels with the toxicity curve Namely, increasing to MTD also means increasing efficacy. We constructed one profile with flat efficacy increase around the MTD (@fig-chemo-flat) and another with steep efficacy increase around the MTD (@fig-chemo-steep).

```{r,echo=FALSE}
#| label: fig-chemo-flat
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a chemo-like profile with a flat efficacy increase around the MTD, the MTD is dose 5."
# see the drug profile
drug_family <- "chemo_flat_efficacy"
profile_parameter <- profile_parameter_list[[drug_family]]
plot_chemo_flat <- plot_drug_profiles(type="chemo",
                   dose_levels =dose_levels,
                   k_tox=profile_parameter$k_tox,L_tox=profile_parameter$L_tox,
                   k_rsp = profile_parameter$k_rsp,L_rsp =profile_parameter$L_rsp,
                   k_surv=profile_parameter$k_surv,L_surv=profile_parameter$L_surv,
                   scale_surv=profile_parameter$scale_surv,
                   offset_surv=profile_parameter$offset_surv,
                   plot_title = "Chemo-like Profile, flat efficacy")
plot_chemo_flat
```

```{r,echo=FALSE}
#| label: fig-chemo-steep
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a chemo-like profile with a steep efficacy increase around the MTD, the MTD is dose 5."
# see the drug profile
drug_family <- "chemo_steep_efficacy"
profile_parameter <- profile_parameter_list[[drug_family]]
plot_chemo_steep <- plot_drug_profiles(type="chemo",
                   dose_levels =dose_levels,
                   k_tox=profile_parameter$k_tox,L_tox=profile_parameter$L_tox,
                   k_rsp = profile_parameter$k_rsp,L_rsp =profile_parameter$L_rsp,
                   k_surv=profile_parameter$k_surv,L_surv=profile_parameter$L_surv,
                   scale_surv=profile_parameter$scale_surv,
                   offset_surv=profile_parameter$offset_surv,
                   plot_title = "Chemo-like Profile, steep efficacy")
plot_chemo_steep
```

```{r,echo=FALSE}
plot_4profiles <- ggarrange(plotlist = list(plot_immuno_OBD3,plot_immuno_OBD4,
                           plot_chemo_flat,plot_chemo_steep),ncol=2,nrow=2)
plot_4profiles
ggsave(filename = paste0("C:\\Users\\ylong\\OneDrive - LUMC\\Projects\\Impacts_early_phase_trial_methods\\plots\\drug_profiles.png"),
       plot = plot_4profiles,
       dpi=320,
       width = 7,
       height = 5,
       units = "in")
```

## Generate individual data for each profile

```{r,echo=FALSE,eval=FALSE}
set.seed(20260303)
ndose <- 6
ndrugs <- 1000
drug_families <- c("immuno_early_plateau", "immuno_late_plateau",
                   "chemo_flat_efficacy", "chemo_steep_efficacy")
# generate individual trial data under this drug profile
for(drug_family in drug_families)
{
# extract corresponding simulation parameters
profile_parameter <- profile_parameter_list[[drug_family]]
# store simulated data
drug_data <- vector(mode = "list",
                    length = ndrugs)
N_ph1 <- 30
N_ph2 <- 300
N_ph3 <- 500
DLT_rates <- scaled_logistic_curve(dose_levels,
                                     k=profile_parameter$k_tox,
                                     L=profile_parameter$L_tox)
if(grepl("immuno",drug_family))
{
rsp_rates <- sigmoid_curve(dose_levels,
                           a= profile_parameter$a_rsp,
                           b=profile_parameter$b_rsp,
                           L=profile_parameter$L_rsp)
median_OS <- sigmoid_curve(dose_levels,
                           a = profile_parameter$a_surv, 
                           b = profile_parameter$b_surv, 
                           L = profile_parameter$L_surv,
                           scale = profile_parameter$scale_surv,
                           offset = profile_parameter$offset_surv)
}
if(grepl("chemo",drug_family))
{
rsp_rates <- scaled_logistic_curve(dose_levels,
                                      k = profile_parameter$k_rsp,
                                      L=profile_parameter$L_rsp)
median_OS <- scaled_logistic_curve(dose_levels,
                                       k=profile_parameter$k_surv,
                                       L=profile_parameter$L_surv,
                                       scale = profile_parameter$scale_surv,
                                       offset = profile_parameter$offset_surv)
}
# simulate 1000 trial data from this profile
for(i in 1:ndrugs)
{
  data_ph1_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph1)[[1]]
  data_ph2_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph2)[[1]]
  data_ph3_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph3)[[1]]
  
  # Response data and survival data
  data_rsp <- gen_rsp(nsim=1,
                      rsp_rates = rsp_rates,
                      N = N_ph2)[[1]]
  
  data_surv <- gen_surv(
    nsim = 1,
    N = N_ph3,
    median_surv0 = 14,
    median_surv1 = median_OS,
    p_trt = 0.5,
    maxt = 60
  )[[1]]
  profile_info <- list(
    drug_id = i,
    dose = dose_levels,
    DLT_rate = DLT_rates,
    rsp_rate = rsp_rates,
    median_OS = median_OS
  )
  drug_data[[i]] <- list("drug_family"=drug_family,
                         "profile_info"=profile_info,
                         "data_ph1_DLT" = data_ph1_DLT,
                         "data_ph2_DLT" = data_ph2_DLT,
                         "data_ph3_DLT" = data_ph3_DLT,
                         "data_rsp" = data_rsp,
                         "data_surv" = data_surv)
if (i%%100 == 0) print(i)
}
save(drug_data,
     file = paste0("simulated_data/",drug_family,".RData"))
}
```
