---
title: "Generate fixed drug profiles for simulation"
author: "Yongxi Long"
format: 
  html:
    self-contained: true
editor: visual
---


## Aim

This is to generate drug profiles of different family:

-   Chemotherapy-like profiles: the toxicity curves and efficacy curves parallel
-   Immunotherapy-lie profiles: the efficacy curves plateau after some dose while the toxicity curves are monotonnically increasing.

Here we generate data from a fixed profile, i.e., all iterations of the simulation evaluate the same drug profile

```{r,warning=FALSE,echo=FALSE}
rm(list = ls())
suppressPackageStartupMessages(
  {
    library(ggplot2)
    library(dplyr)
    library(ggpubr)
    library(simsurv)
  }
)
source("data_generating_functions.R")
```

## Function skeletons

1.  For toxicity, we use more a monotonic function, the scaled logistic:

$$P(Y=1\mid d) = L \frac{1-e^{-kd}}{1-e^{-k}}$$

-   $k$ adjusts the curve growth rate.

2.  For response and survival, we use an S-shaped skeleton to obtain a plateau:

$$P(Y=1\mid d) = L \frac{d^a}{d^a+(1-d)^b}$$

a and b are shape control parameters:

-   if a=b, the S-curve is symmetric around d = 0.5

-   if $a<b$, the S-curve rises faster initially, otherwise the curve starts very flat and rises more at the end.

-   For dose-survival, we first obtain some value between 0 and 1 from the above function, then scale them by some constant to have proper units (e.g., days, months, years). We also add an offset to make sure that the survival time at dose zero is a positive value (otherwise it will be zero).

## Immuno-like drug profiles

In this type of drug profiles (immuno/targeted therapies), the efficacy curves plateau before MTD. Namely, the optimal biological dose (OBD) is smaller than the maximum tolerable dose (@fig-dose-tox-rsp-surv). We set $a<b$ for dose-response, dose-survival curves to achieve this. For dose-(median) survival curve, we add scale and offset parameter in the function to make the survival time in proper range and has a minimum positive value.

How earlier the efficacy plateau is relative to the MTD is a tuning parameter. We'll evaluate the case where OBD is much separated from MTD (@fig-immuno-OBD3-MTD5), as well as where the OBD and the MTD is close (@fig-immuno-OBD4-MTD5).

```{r,echo=FALSE}
#| label: fig-immuno-OBD3-MTD5
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a immunotherapy-like profile, the OBD is dose 3, which is very separated from MTD which is dose 5."
# get dose levels
# dose selection
ndose <- 6
lowest_dose <- 0.1
dose_levels <- lowest_dose*c(1,2,2*1.65,2*1.65*1.52,2*1.65*1.52*1.4,2*1.65*1.52*1.4*1.33)
# parameters for the curves
k_tox = -3
L_tox = 0.8
L_rsp=0.5
L_surv = 0.6
a_rsp <- 1
b_rsp <- 12
a_surv <- 0.1
b_surv <- 12
scale_surv <- 20
offset_surv <- 8

# see the drug profile
plot_drug_profiles(dose_levels = dose_levels,
                   k_tox=k_tox,L_tox=L_tox,
                   a_rsp=a_rsp,b_rsp=b_rsp,L_rsp=L_rsp,
                   a_surv=a_surv,b_surv=b_surv,L_surv=L_surv,
                   scale_surv=scale_surv,offset_surv=offset_surv)
```

```{r,echo=FALSE,eval=FALSE}
# generate individual trial data under this drug profile
drug_family <- "immuno_fixed_OBD3_MTD5"
set.seed(1006)
ndose <- 6
ndrugs <- 1000
# store simulated data
drug_data <- vector(mode = "list",
                    length = ndrugs)
N_ph1 <- 30
N_ph2 <- 300
N_ph3 <- 500
DLT_rates <- scaled_logistic_curve(dose_levels,
                                     k=k,
                                     L=L_tox)
rsp_rates <- sigmoid_curve(dose_levels, a= a_rsp,b=b_rsp,L=L_rsp,
  )
median_OS <- sigmoid_curve(dose_levels,
                             a = a_surv, b = b_surv, L = L_surv,
                             scale = scale_surv,
                             offset = offset_surv)
parameter <- c("k"=k,"L_tox"=L_tox,
               "a_rsp"=a_rsp,"b_rsp"=b_rsp,"L_rsp"=L_rsp,
               "a_surv" = a_surv,"b_surv" = b_surv,"L_surv"=L_surv)
# simulate 1000 trial data from this one fixed profile
for(i in 1:ndrugs)
{
  data_ph1_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph1)[[1]]
  data_ph2_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph2)[[1]]
  data_ph3_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph3)[[1]]
  
  # Response data and survival data
  data_rsp <- gen_rsp(nsim=1,
                      rsp_rates = rsp_rates,
                      N = N_ph2)[[1]]
  
  data_surv <- gen_surv(
    nsim = 1,
    N = N_ph3,
    median_surv0 = 14,
    median_surv1 = median_OS,
    p_trt = 0.5,
    maxt = 60
  )[[1]]
  profile_info <- list(
    drug_id = i,
    dose = dose_levels,
    DLT_rate = DLT_rates,
    rsp_rate = rsp_rates,
    median_OS = median_OS,
    parameter = parameter
  )
  drug_data[[i]] <- list("drug_family"=drug_family,
                         "profile_info"=profile_info,
                         "data_ph1_DLT" = data_ph1_DLT,
                         "data_ph2_DLT" = data_ph2_DLT,
                         "data_ph3_DLT" = data_ph3_DLT,
                         "data_rsp" = data_rsp,
                         "data_surv" = data_surv)
if (i%%100 == 0) print(i)
}
save(drug_data,
     file = paste0("simulated_data/",drug_family,".RData"))
```

```{r,echo=FALSE}
#| label: fig-immuno-OBD4-MTD5
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a immunotherapy-like profile, the OBD is dose 4, which is close to MTD which is dose 5."
# get dose levels
# dose selection
ndose <- 6
lowest_dose <- 0.1
dose_levels <- lowest_dose*c(1,2,2*1.65,2*1.65*1.52,
                             2*1.65*1.52*1.4,2*1.65*1.52*1.4*1.33)
# parameters for the curves
k_tox = -3
L_tox = 0.8
L_rsp=0.5
L_surv = 0.6
a_rsp <- 2
b_rsp <- 8
a_surv <- 2
b_surv <- 8
scale_surv <- 20
offset_surv <- 8

# see the drug profile
plot_drug_profiles(dose_levels = dose_levels,
                   k_tox=k_tox,L_tox=L_tox,
                   a_rsp=a_rsp,b_rsp=b_rsp,L_rsp=L_rsp,
                   a_surv=a_surv,b_surv=b_surv,L_surv=L_surv,
                   scale_surv=scale_surv,offset_surv=offset_surv)
```

```{r,echo=FALSE,eval=FALSE}
# generate individual trial data under this drug profile
drug_family <- "immuno_fixed_OBD4_MTD5"
set.seed(1006)
ndose <- 6
ndrugs <- 1000
# store simulated data
drug_data <- vector(mode = "list",
                    length = ndrugs)
N_ph1 <- 30
N_ph2 <- 300
N_ph3 <- 500
DLT_rates <- scaled_logistic_curve(dose_levels,
                                     k=k,
                                     L=L_tox)
rsp_rates <- sigmoid_curve(dose_levels, a= a_rsp,b=b_rsp,L=L_rsp,
  )
median_OS <- sigmoid_curve(dose_levels,
                             a = a_surv, b = b_surv, L = L_surv,
                             scale = scale_surv,
                             offset = offset_surv)
parameter <- c("k"=k,"L_tox"=L_tox,
               "a_rsp"=a_rsp,"b_rsp"=b_rsp,"L_rsp"=L_rsp,
               "a_surv" = a_surv,"b_surv" = b_surv,"L_surv"=L_surv)
# simulate 1000 trial data from this one fixed profile
for(i in 1:ndrugs)
{
  data_ph1_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph1)[[1]]
  data_ph2_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph2)[[1]]
  data_ph3_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph3)[[1]]
  
  # Response data and survival data
  data_rsp <- gen_rsp(nsim=1,
                      rsp_rates = rsp_rates,
                      N = N_ph2)[[1]]
  
  data_surv <- gen_surv(
    nsim = 1,
    N = N_ph3,
    median_surv0 = 14,
    median_surv1 = median_OS,
    p_trt = 0.5,
    maxt = 60
  )[[1]]
  profile_info <- list(
    drug_id = i,
    dose = dose_levels,
    DLT_rate = DLT_rates,
    rsp_rate = rsp_rates,
    median_OS = median_OS,
    parameter = parameter
  )
  drug_data[[i]] <- list("drug_family"=drug_family,
                         "profile_info"=profile_info,
                         "data_ph1_DLT" = data_ph1_DLT,
                         "data_ph2_DLT" = data_ph2_DLT,
                         "data_ph3_DLT" = data_ph3_DLT,
                         "data_rsp" = data_rsp,
                         "data_surv" = data_surv)
if (i%%100 == 0) print(i)
}
save(drug_data,
     file = paste0("simulated_data/",drug_family,".RData"))
```

## Chemo-like drug profiles

In this type of drug profiles (chemo therapies), the efficacy curves parallels with the toxicity curve Namely, increasing to MTD also means increasing efficacy (@fig-chemo-MTD5)

```{r}
#| label: fig-chemo-MTD5
#| fig-cap: "Dose-DLT, dose-response, and dose-median OS for a chemo-like profile, the MTD is dose 5."
# get dose levels
# dose selection
ndose <- 6
lowest_dose <- 0.1
dose_levels <- lowest_dose*c(1,2,2*1.65,2*1.65*1.52,2*1.65*1.52*1.4,2*1.65*1.52*1.4*1.33)
# parameters for the curves
k_tox = -3
L_tox = 0.8
L_rsp=0.58
L_surv = 0.68
# a_rsp <- 10
# b_rsp <- 6
# a_surv <- 5
# b_surv <- 3
k_rsp <- 2
k_surv <- 2
scale_surv <- 20
offset_surv <- 8

# see the drug profile
plot_drug_profiles(type="chemo",
                   dose_levels =dose_levels,
                   k_tox=k_tox,L_tox=L_tox,
                   k_rsp = k_rsp,L_rsp =L_rsp,
                   k_surv=k_surv,L_surv=L_surv,
                   scale_surv=scale_surv,
                   offset_surv=offset_surv)
```


```{r,echo=FALSE,eval=FALSE}
# generate individual trial data under this drug profile
drug_family <- "chemo_fixed_MTD5"
set.seed(1006)
ndose <- 6
ndrugs <- 1000
# store simulated data
drug_data <- vector(mode = "list",
                    length = ndrugs)
N_ph1 <- 30
N_ph2 <- 300
N_ph3 <- 500
DLT_rates <- scaled_logistic_curve(dose_levels,
                                     k=k_tox,
                                     L=L_tox)
rsp_rates <- scaled_logistic_curve(dose_levels,
                                      k = k_rsp,
                                      L=L_rsp)
median_OS <- scaled_logistic_curve(dose_levels,
                                       k=k_surv,
                                       L=L_surv,
                                       scale = scale_surv,
                                       offset = offset_surv)
parameter <- c("k_tox"=k_tox,"L_tox"=L_tox,
               "k_rsp" = k_rsp,"L_rsp"=L_rsp,
               "k_surv" = k_surv,"L_surv"=L_surv)
# simulate 1000 trial data from this one fixed profile
for(i in 1:ndrugs)
{
  data_ph1_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph1)[[1]]
  data_ph2_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph2)[[1]]
  data_ph3_DLT <- gen_DLT(nsim=1,
                          DLT_rates = DLT_rates,
                          N = N_ph3)[[1]]
  
  # Response data and survival data
  data_rsp <- gen_rsp(nsim=1,
                      rsp_rates = rsp_rates,
                      N = N_ph2)[[1]]
  
  data_surv <- gen_surv(
    nsim = 1,
    N = N_ph3,
    median_surv0 = 14,
    median_surv1 = median_OS,
    p_trt = 0.5,
    maxt = 60
  )[[1]]
  profile_info <- list(
    drug_id = i,
    dose = dose_levels,
    DLT_rate = DLT_rates,
    rsp_rate = rsp_rates,
    median_OS = median_OS,
    parameter = parameter
  )
  drug_data[[i]] <- list("drug_family"=drug_family,
                         "profile_info"=profile_info,
                         "data_ph1_DLT" = data_ph1_DLT,
                         "data_ph2_DLT" = data_ph2_DLT,
                         "data_ph3_DLT" = data_ph3_DLT,
                         "data_rsp" = data_rsp,
                         "data_surv" = data_surv)
if (i%%100 == 0) print(i)
}
save(drug_data,
     file = paste0("simulated_data/",drug_family,".RData"))
```